---
title: "Exploratory data analysis of data by offense type"
author: "Riccardo Fogliato"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EDA by offense type}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  offense: '13B'
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(rmarkdown.html_vignette.check_title = FALSE)
```

## Setup: packages and data

```{r load_packages, include=FALSE,  warning = FALSE, message = FALSE}
library(tidyverse)
library(vroom)
library(here)
library(knitr)
library(kableExtra)
library(xtable)
library(quantreg)
library(broom)

source(here('analysis', 'utils.R'))
```


```{r setup_plots, include = FALSE}
list_states <- get_country_div_and_state()
states_order <- list_states$states_order
country_div_order <- list_states$country_div_order
states_order <- rev(states_order); country_div_order <- rev(country_div_order)
country_list_color <- sapply(country_div_order, get_color_division)
levels_pop_group <- c("Cities < 2,500" , "Cities 2,500-9,999", "Cities 10,000-24,999", 
                      "Cities 25,000-49,999", "Cities 50,000-99,999", 
                      "Cities 100,000-249,999", 
                      "Cities 250,000-499,999","Cities 500,000-999,999", 
                      "MSA Counties < 10,000", "MSA Counties 10,000-24,999", 
                      "MSA Counties 25,000-99,999",
                      "MSA Counties 100,000+","MSA State Police" , 
                      "Non-MSA Counties < 10,000" , "Non-MSA Counties 10,000-24,999", 
                      "Non-MSA Counties 25,000-99,999", "Non-MSA Counties 100,000+", 
                      "Non-MSA State Police")
```


```{r, include = FALSE, eval = FALSE}
# if you run this from rstudio, then you should run this line
offense <- '13B'
```

```{r, include = FALSE}
offense <- params$offense
```


The offense in this report is `offense`.
```{r load_data, include = TRUE,  warning = FALSE}
df <- vroom(here('data', paste0('data_', offense, '.csv')),
            col_types = cols())
```

```{r rename_again_murder, include = FALSE}
offense <- rename_offenses(offense); offense
offense <- ifelse(offense == "Murder/non-negligent manslaughter", 'Murder non-negligent manslaughter', offense)
```


## Goal 

In this analysis, we focus the rest of the analysis of variations in arrest rates across jurisdictions. 
Again, this analysis is relative to ``offense``.


## Analysis

Consider only incidents that have not been cleared by exceptional means.
```{r exclude_excep, include = TRUE,  warning = FALSE, message = FALSE}
df <- df %>% filter(cleared_exceptionally == "not applicable")
```

```{r check_race_dist, include = FALSE, warning = FALSE, message = FALSE}
df %>% count(race_of_offender) %>% mutate(prop = n/sum(n))
df %>% group_by(race_of_offender, race_of_victim) %>% summarise(n = n()) %>% mutate(prop = n/sum(n))
```

Get the overall arrest rate (y=1 means that an arrest happened).
```{r arrest_rate_race,  warning = FALSE, message = FALSE}
df %>% group_by(race_of_offender) %>% 
  summarise(n_offenses = n(), 
            arrest_rate = round(mean(y),2)) %>%
  sable()
```



Let's look at the arrest rate across states. 
```{r arrest_rate_by_state,  warning = FALSE, message = FALSE}
arrest_by_state <- df %>% 
  group_by(state_abbv) %>% 
  summarise(
    `Arrest rate` = round(mean(y),2),
    `Arrest rate white-black` = round(mean(y[race_of_offender=='White'])-mean(y[race_of_offender=='Black']),2), 
    `Nb black offenders` = sum(race_of_offender=='Black'), 
    `Nb white offenders` = sum(race_of_offender=='White'))

arrest_by_state %>%
    sable()
```

Let's compute the mean and standard deviation of the difference in arrest rates of white-black across states.
```{r arrest_rate_by_state_std.dev,  warning = FALSE, message = FALSE}
arrest_by_state %>%
  summarise(
    `Mean diff. in arrest rates white-black` = round(mean(`Arrest rate white-black`), 2),
    `Standard deviation of diff. in arrest rates white-black` = round(sd(`Arrest rate white-black`), 2)) %>%
      sable()
```

Let's also plot these differences. 
```{r arrest_rate_by_state_plot, include = TRUE,  warning = FALSE, fig.height = 7, fig.width = 7}
p <- df %>%
  group_by(state_abbv, race_of_offender) %>%
  summarise(y_mean = mean(y), std.err = sd(y)/sqrt(n()), n = n()) %>%
  filter(n > 50) %>%
  ggplot(aes(factor(state_abbv, levels=states_order), y_mean, 
             fill = race_of_offender)) + 
  geom_col(position = position_dodge()) + 
  geom_errorbar(aes(ymin = y_mean-1.96*std.err, ymax = y_mean+1.96*std.err), 
                width = 0.2, position=position_dodge(.9)) +
  xlab('State code') +  ylab('Arrest rate') + theme_bw() +
  scale_fill_manual('Race of offender', values = c('grey60', "#E69F00")) + 
  theme(axis.text.x = 
          element_text(colour = country_list_color)) + 
  ylim(0,1) + 
  ggtitle(paste0('Arrest rates for ', tolower(offense)))
p
```


```{r arrest_rate_by_state_plot_save, include = FALSE,  warning = FALSE}
p <- p  +
  theme(plot.title = element_text(size=40), axis.title = element_text(size=30), 
        axis.text = element_text(size = 20), legend.text = element_text(size=20),
        legend.title = element_text(size = 20))
ggsave(width = 14, height = 10, 
       here('figures', paste0('Arrest_rate_by_state_', offense, '.pdf')))
```


## Focus on TN, MI, SC

Given the large differences in arrest rates observed in Tennesse, Michigan, and South Carolina, let's focus on these states.

Let's keep only incidents from these states. 
```{r subset_states_high,  warning = FALSE, message = FALSE}
df_state <- df %>% filter(grepl('SC|MI|TN', state_abbv))
```

```{r rename_nashville, include = FALSE}
df_state <- df_state %>%
  mutate(city_name = ifelse(city_name == 'NASHVILLE METROPOLITAN', 'NASHVILLE', city_name))
```


Let's first look at Tennessee and Michigan, which both contain cities with > 500k citizens as recorded in the data.
We now compute the arrest rates in those cities vs the rest. 
```{r states_high_arrest_rate,  warning = FALSE, message = FALSE}
df_state %>% filter(grepl('MI|TN', state_abbv)) %>% 
  # SC does not have cities > 500k
  mutate(`Large city` = ifelse(population_group == 'Cities 500,000-999,999', city_name, 'No')) %>%
  group_by(state_abbv, `Large city`) %>% 
  summarise(
            `Number of offenses` = n(),
            `Arrest rate` = round(mean(y),2) ,
            `Arrest rate black` = round(mean(y[race_of_offender == "Black"]),2),
            `Arrest rate white` = round(mean(y[race_of_offender == "White"]),2),
            `Diff. arrest rate white-black` = round(mean(y[race_of_offender=='White'])-mean(y[race_of_offender=='Black']),2),
            `% of offenders that is black in city` = round(mean(race_of_offender=='Black'),2)) %>%
        sable()
```


Given the large differences in the arrest rates across cities, we check whether the distribution of the population in large cities vs. other areas can explain the overall gap.
```{r distribution_offenders_cities, warning = FALSE, message = FALSE}
df_state %>% filter(grepl('MI|TN', state_abbv)) %>% 
  mutate(`Large city` = ifelse(population_group == 'Cities 500,000-999,999', city_name, 'No')) %>%
  group_by(state_abbv, race_of_offender, `Large city`) %>%
  summarise(`Number of offenders` = n()) %>%
  mutate(`% of offenders` = round(`Number of offenders`/sum(`Number of offenders`),2)) %>%
  sable()
```


We now investigate the variation in arrest rates across core cities.
```{r core_city_arrest_rate, warning = FALSE, message = FALSE}
core_city_arrest_rate <- df_state %>%
  filter(core_city == 'Yes') %>%
  group_by(state_abbv, city_name) %>% 
  summarise(
            `Number of offenses` = n(),
            `Arrest rate` = round(mean(y),2) ,
            `Arrest rate black` = round(mean(y[race_of_offender == "Black"]),2),
            `Arrest rate white` = round(mean(y[race_of_offender == "White"]),2),
            `Diff. arrest rate white-black` = round(mean(y[race_of_offender=='White'])-mean(y[race_of_offender=='Black']),2),
            `% of offenders that is black in city` = round(mean(race_of_offender=='Black'),2))

core_city_arrest_rate %>%
        sable()
```

We then check whether arrest rates are lower in core cities compared to the rest of the state. 
```{r arrest_rate_core_vs_rest, warning = FALSE, message = FALSE}
df_state %>%
  group_by(state_abbv, core_city) %>% 
  summarise(
            `Number of offenses` = n(),
            `Arrest rate` = round(mean(y),2) ,
            `Arrest rate black` = round(mean(y[race_of_offender == "Black"]),2),
            `Arrest rate white` = round(mean(y[race_of_offender == "White"]),2),
            `Diff. arrest rate white-black` = round(mean(y[race_of_offender=='White'])-mean(y[race_of_offender=='Black']),2),
            `% of offenders that is black in city` = round(mean(race_of_offender=='Black'),2)) %>%
  sable()
```


## Analysis of arrest rates across police agencies (ORIs)

We first create summary statistics for each police agency. 
```{r get_ori_stats, warning = FALSE, message = FALSE}

ori_stats <- df %>%
  group_by(originating_agency_identifier) %>%
  summarise(
    `Nb offenses` = n(),
    `Arrest rate` = mean(y),
  `Diff. arrest rate white-black` = sum(y==1 & race_of_offender=='White')/sum(race_of_offender=='White')-sum(y==1 & race_of_offender=='Black')/sum(race_of_offender=='Black'),
    `Arrest rate black` = sum(y==1 & race_of_offender=='Black')/sum(race_of_offender=='Black'),
    `Arrest rate white` = sum(y==1 & race_of_offender=='White')/sum(race_of_offender=='White'),
    `Nb black offenders` = sum(race_of_offender=='Black'),
    `Nb white offenders` = sum(race_of_offender=='White'),
    `% offenders that is black` = sum(race_of_offender=='Black')/n(),
    # population served & officers characteristics
    `Population served` = mean(population_leoka, na.rm=T),
    `Officer rate` = mean(officer_rate, na.rm=T),
    `Population group` = population_group[1],
    `State` = state_abbv[1],
    `Total officers` = mean(total_off, na.rm=T),
    `Core city` = mean(core_city=='Yes'),
  `Crime per capita` = n()/`Population served`
  )
glimpse(ori_stats)
```

Importantly, we keep only agencies with at least 100 crimes recorded in the period considered (i.e., on average approx. 10 crimes a year). The omission will likely exclude agencies in rural areas. 
```{r subset_oris, warning = FALSE, message = FALSE}
length(unique(ori_stats$originating_agency_identifier))
ori_stats <- ori_stats %>%
  # include those with > 100 crimes
  # !!! this affects subsequent results !!!
  filter(`Nb offenses` > 100) 
length(unique(ori_stats$originating_agency_identifier))
```


```{r, include = FALSE}
# number of police agencies
length(unique(ori_stats$originating_agency_identifier))
```


```{r, include = FALSE}
ori_stats %>% ggplot(aes(`Nb offenses`)) + geom_histogram() + xlim(0,5000)
```


Let's plot the gap in arrest rates between white and black offenders for all jurisdictions with at least 50 offenses with Black offenders and 50 offenses with White offenders.
```{r gap_arrest_rate_plot, warning = FALSE, message = FALSE, fig.height = 7, fig.width = 7}
# gap in arrest rates by police agency -- order states
state_gap <- df %>% ungroup %>%
  group_by(state_abbv) %>% 
  summarise(y_diff = mean(y[race_of_offender=='White'])-mean(y[race_of_offender=='Black']))

p <- ori_stats %>%
  filter(`Nb black offenders`>50 & `Nb white offenders`>50) %>%
  ggplot(aes(factor(`State`, levels = states_order), `Diff. arrest rate white-black`)) + 
  geom_boxplot() + 
  theme_bw() + 
  ylab(expression(atop('Difference in arrest rates between', 'white and black offenders'))) + 
  xlab('State code') + 
  geom_hline(yintercept=0, linetype = 'dashed') + 
  geom_point(data=state_gap, aes(state_abbv, y_diff), shape = 8, size = 5, col = 'forestgreen') + 
  ylim(-0.15,0.21) + 
  theme(axis.text.x = 
          element_text(colour = country_list_color))
p
```


```{r gap_arrest_rate_plot_save, include = FALSE}
p + theme(text = element_text(size=40), axis.text = element_text(size = 30))
ggsave(width = 20, height = 15, here('figures', paste0('Arrest_rate_by_pol_agency_', offense, '2.pdf')))
p  + 
  theme(text = element_text(size=25), axis.text = element_text(size = 25))
ggsave(width = 20, height = 8, here('figures', paste0('Arrest_rate_by_pol_agency_', offense, '.pdf')))
```


Let's then focus on the relationship that we observed for large cities. 
We compute some correlations between the share of black offenders in the police agency and other characteristics of the agency.

```{r cor_off_crimecapita}
cor.test(ori_stats$`% offenders that is black`, ori_stats$`Crime per capita`)
```


```{r cor_off_officerrate}
cor.test(ori_stats$`% offenders that is black`, ori_stats$`Officer rate`)
```

```{r cor_off_popserved}
cor.test(ori_stats$`% offenders that is black`, ori_stats$`Population served`)
```

```{r cor_off_officerperinc}
cor.test(ori_stats$`% offenders that is black`, ori_stats$`Total officers`/ori_stats$`Nb offenses`)
```

```{r cor_off_arrestrate}
cor.test(ori_stats$`% offenders that is black`, ori_stats$`Arrest rate`)
```

```{r additional_checks, include = FALSE}
rq(ori_stats$`Crime per capita` ~ ori_stats$`% offenders that is black`) %>% tidy()
rq(ori_stats$`Officer rate` ~ ori_stats$`% offenders that is black`) %>% tidy()
rq(ori_stats$`Population served` ~ ori_stats$`% offenders that is black`) %>% tidy()
rq(ori_stats$`Total officers`/ori_stats$`Nb offenses` ~ ori_stats$`% offenders that is black`) %>% tidy()
rq(ori_stats$`Arrest rate` ~ ori_stats$`% offenders that is black`) %>% tidy()
```

We now plot the percentage of black offenders as a function of the arrest rate in the police agency.
```{r arrest_rate_vs_black_prop, warning = FALSE, message = FALSE, fig.height = 10, fig.width = 10}
p <- ori_stats %>% ungroup %>%
  pivot_longer(cols = c(`Nb white offenders`, `Nb black offenders`), 
               names_to = 'Race of offenders', 
               values_to = 'Nb offenders') %>%
  mutate(`Arrest rate binned` = floor(pmin(`Arrest rate`,0.99)*10)/10) %>%
  group_by(`State`, `Race of offenders`, `Arrest rate binned`) %>%
  summarise(`Total nb offenders` = sum(`Nb offenders`)) %>% 
  mutate(`% offenders` = `Total nb offenders`/sum(`Total nb offenders`)) %>%
  ggplot(aes(`Arrest rate binned`, `% offenders`, fill = `Race of offenders`)) + 
  geom_col(position = position_dodge2()) + theme_bw() + 
  xlab('Arrest rate in police agency') + 
  facet_wrap( ~ `State`, ncol = 2) + 
  ylab('Density') + 
  scale_fill_manual('Race of offender', 
                    values = c('grey60', "#E69F00"), 
                    labels = c('Black', 'White')) + 
  scale_x_continuous(breaks = seq(0,.9,by=0.1), 
                     labels = paste0(seq(0,90,by=10), '-', seq(10,100,by=10), '%'))

p #+ theme(plot.title = element_text(size=40), axis.title = element_text(size=20))
```

```{r arrest_rate_vs_black_prop_save, include = FALSE}
p + theme(plot.title = element_text(size=40), axis.title = element_text(size=20))
ggsave(height = 20, width = 14, 
       here('figures', paste0('Distribution_by_arrest_rate_', offense, '.pdf')))
```


```{r pop_served_vs_officer_rate, include = FALSE, warning = FALSE, message = FALSE}
ori_stats %>%
  ggplot(aes(`Population served`, `Officer rate`)) + geom_point() + 
  xlim(0,30000) + ylim(0,50) + theme_bw() + 
  xlab('Population served by police agency') + 
  ylab('Officers per 1000 capita in police agency')
cor.test(ori_stats$`Population served`, ori_stats$`Officer rate`)
```

```{r black_pop_vs_arrestrate, include = FALSE, fig.width = 7, fig.height = 10}
# difference in arrest rates as a function on black pop
ori_stats %>% ggplot(aes(`% offenders that is black`, `Diff. arrest rate white-black`)) +
  geom_point() + ylim(-1,1) + 
  ylab('Difference in arrest rates between White and Black offenders') + 
  xlab('% offenders that are Black in police agency')
```

