---
title: "Exploratory data analysis of full data"
author: "Riccardo Fogliato"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EDA of full data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(rmarkdown.html_vignette.check_title = FALSE)
```

## Setup: packages and data

```{r load_packages, include=FALSE}
library(tidyverse)
library(vroom)
library(here)
library(xtable)
library(knitr)
library(kableExtra)

source(here('analysis', 'utils.R'))
```

```{r load_data, include = TRUE, warning = FALSE}
admin_batch <- vroom(here('data', 'admin_batch.csv'),
                     col_types = cols())
offense <- vroom(here('data', 'offense.csv'),
                 col_types = cols())
```

```{r subset_data_offense, include = TRUE, warning = FALSE}
# subset offenses of interest
offense_sub <- offense %>% 
  distinct(year, incident_number, 
           ucr_offense_code, originating_agency_identifier) %>% 
  filter(grepl('13A|13B|Aggravated|Simple|Murder|09A|11A|Forcible Rape|Robbery|120|(13A)|(13B)|(120)|(09A)|(11A)',
               ucr_offense_code))

# rename offenses
offense_sub <- offense_sub %>%
  mutate(offense = case_when(
    grepl('09A|Murder', ucr_offense_code) ~ '09A',
    grepl('13A|Aggravated', ucr_offense_code) ~ '13A',
    grepl('13B|Simple', ucr_offense_code) ~ '13B',
    grepl('120|Robbery|120', ucr_offense_code) ~ '120',
    grepl('11A|Forcible', ucr_offense_code) ~ '11A'))
```


```{r subset_states, include = TRUE, warning = FALSE}
# subset states of interest
states <- 'AR|CO|DE|ID|IA|KY|MI|MT|NH|ND|SC|SD|TN|VT|VA|WV'
admin_batch <- admin_batch %>% filter(grepl(states, state_abbv))
```


```{r create_df, include = TRUE, warning = FALSE}
# join data
df <- admin_batch %>% 
  distinct(year, incident_number, 
           cleared_exceptionally, originating_agency_identifier, 
           total_arrestee_segments) %>%
  inner_join(offense_sub, 
             by = c("originating_agency_identifier", "incident_number", "year"))
```

## Goal
In this file, we look at clearance and arrest rates for all incidents with at 
least one violent offense. 


## Analysis

First, let's compute the yearly number of offenses.
```{r crimes_year, warning = FALSE}
# number of crimes per year
df %>% count(offense, year) %>% sable()
```

We now look at the number of incidents and clearance rates (i.e., incidents that include at least an exceptional clearance or one arrest).
```{r clearance_rate, warning = FALSE}
# compute arrest rates (with and without exceptional clearance)
n_per_offense <- df %>% group_by(offense) %>% summarise(n = n()) %>%
  pivot_wider(names_from = offense, values_from = n)
clearance_rate <- df %>% group_by(offense) %>% 
  summarise(arrest_rate = round(mean(ifelse(total_arrestee_segments > 0 | 
                      cleared_exceptionally != 'not applicable', 1, 0)),2)) %>%
  pivot_wider(names_from = offense, values_from = arrest_rate)

summary_stats <- n_per_offense %>%
      bind_rows(clearance_rate)
summary_stats %>% sable()
```


```{r clearance_rate_save, include = FALSE, warning = FALSE}
summary_stats %>%
  xtable(., type = "latex")
```


Let's compute the share of incidents that result in at least one arrest.
But let's first drop those that are cleared by exceptional means.
```{r arrest_rate, warning = FALSE}
n_per_offense_wo_excep <- df %>% filter(cleared_exceptionally == 'not applicable') %>% 
  group_by(offense) %>% summarise(n = n()) %>% 
  pivot_wider(names_from = offense, values_from = n)
arrest_rate <- df %>% filter(cleared_exceptionally == 'not applicable') %>% 
  group_by(offense) %>% summarise(arrest_rate = round(mean(mean(total_arrestee_segments>0)),2)) %>%
    pivot_wider(names_from = offense, values_from = arrest_rate)

summary_stats <- n_per_offense %>%
      bind_rows(arrest_rate)
summary_stats %>% sable()
```

```{r arrest_rate_save, include = FALSE}
#summary_stats %>%
#  xtable(., type = "latex")
```

